PyTracer: Automatically profiling numerical instabilities in Python

### 2021-09-21 paper.tex
Tristan's review on sections 3 and 4.1.1

\documentclass[11pt]{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{todonotes}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pifont}
\usepackage{xcolor,colortbl}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{amsmath}

\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}
\newcommand{\Yohan}[1]{\todo[inline,backgroundcolor=green]{YC:#1}}
\newcommand{\cross}[0]{\cellcolor{red!65}\ding{53}}
\newcommand{\valid}[0]{\cellcolor{green!75!black}\ding{51}}
\newcommand{\warn}[0]{\cellcolor{orange!75}?}

@@

@@

@@

The scientific Python ecosystem is a central ingredient of scientific
analyses due to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications every year, providing a reference set of high-quality open-source core scientific libraries. Numerous domain-specific data analysis pipelines were built from this ecosystem by diverse research groups, leveraging Python's simplicity and expressivity.

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, tiny numerical perturbations introduced
by data noise, software and hardware updates, or parallelization lead to
substantial deviations in final results and potentially different scientific conclusions, threatening the reliability of computational analyses. To address this issue, numerical stability analyses need to be conducted systematically. However, no practical tool currently exists to conduct such analyses on Python programs.

We present PyTracer, a numerical stability analyzer for Python programs.
PyTracer adopts a dynamic approach that evaluates numerical stability using program execution traces, allowing scaling to large programs without manual intervention. In contrast, theoretical approaches based on condition number or backward error analyses require detailed modelling of the program and its numerical implementation. Similarly, static code analysis techniques such as Frama-C~\cite{cuoq2012frama}, Gappa~\cite{de2010certifying} or Flocq~\cite{boldo2011flocq} hardly scale to large codebases, particularly in dynamically-typed languages such as Python.

@@

In this section, we review tools and theoretical frameworks related to Pytracer components.
We start by reviewing state-of-the-art dynamic tools for detecting numerical instabilities.
Then, we focus on existing theoretical approaches for perturbations-based techniques to detect numerical instabilities,
Pytracer being agnostic to perturbation technique used. Finally, we review visualization tools 
aiming at detecting numerical instabilities over execution time.

\label{sec:soa}
\subsection{Numerical assessment tools}

Several tools have been and continue to be developed to assess numerical quality.
Those tools can be divided in two main categories: static VS dynamic analysis.
Static analysis focuses on safety and soundness of error bounds provided 
while dynamic analysis rather provides proof-less numerical quality estimations.
The main drawback of static analysis tools is the poor scalability that confine them to small programs.
Pytracer falls into the second category by targeting real Python use cases.
In this section, we will hence review the most known tools based on dynamic analysis.
We divide the tools in three sections starting by tools dedicated to non Python codes, then 
tools using Dynamic Binary Instrumentation and finally tools dedicated to Python codes.

Following the languages mainly used in HPC or embedded systems, 
most the tools are dedicated to C, C++ or FORTRAN codes. 
Those tools are either source-to-source or compiler based transformations.
CADNA~\cite{jezequel2008cadna} is a library for C, C++ and Fortran codes implementing
the CESTAC~\cite{vignes1993stochastic} method, a stochastic arithmetic used for detecting round-off
errors by introducing numerical instabilities in floating-point computations.
Verificarlo~\cite{verificarlo} is a compiler that instruments floating point operations
to replace them by generic call to Verificarlo API, providing several backends implementations including one for 
Monte Carlo Arithmetic (MCA) (cf.~\ref{sec:mca}), another stochastic arithmetic.
Bao et al. ~\cite{bao2013fly} modifies gcc to track local floating-point error across execution.
pLiner~\cite{guo2020pliner} is root cause analyzer based upon clang that detects floating point operations 
responsible of result variability. 
It uses source-to-source transformation at AST level to rewrite parts of code with higher precision and to compare the result with
the original one. PFPSanatizer~\cite{chowdhary2021parallel} is a compiler that uses parallel shadow execution to detect numerical issues
by using higher precision. 
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model to propagate numerical error. 
FLiT~\cite{sawaya2017flit} is framework to detect variability induced by compilers and their optimizations.
Tang et al.~\cite{tang2016software} propose a source-to-source framework for executing targeted code 
in infinite and fixed precision with and without using stochastic arithmetic.

The main other existing approach uses Dynamic Binary Instrumentation (DBI) to perform analysis at the binary level,
removing the language barrier and allowing to analyze any binary without needing the source code. This approach
is particularly relevant on closed-source software. However, working at binary level loses high-level information useful
for debugging and understanding the source code semantic and logic.
CraftHPC~\cite{lam2013dynamic} uses DBI to detect catastrophic cancellations at runtime.
Verrou~\cite{fevotte2016verrou} is a Valgrind~\cite{nethercote2007valgrind} based tool that replaces on-the-fly
floating-point operations by their stochastic arithmetic counterparts. FPDebug~\cite{benz2012dynamic} uses DBI to detect numerical 
inaccuracies by using shadow computations with a higher precision.
Herbgring~\cite{sanchez2017finding} is a Valgrind-based tool to detect
numerical instabilities. It uses a shadow memory to do detect precision losses by comparison with results obtained with the high-precision 
MPFR library~\cite{fousse2007mpfr}. It is combined with symbolic computation to backtrack the root of the error.

Existing tools for tracing Python code are dedicated to performance profiling, for time (cProfile) 
or memory consumption (memprofile). 
To the best of our knowledge, there is currently no tool dedicated to numerical analysis.
We can nevertheless mention Anteater~\cite{faust2019anteater} a Python tracer tool that aims at debugging Python application. 
It operates source transformation at AST level but only deals with native numeric Python types.
However, we can compare the methodology used by these tools against the one used in Pytracer.
These tools use an instrumentation based on decorators, a Python mechanism allowing 
to wrap a function by adding a line over a function declaration.
While this method is useful when targeting specific functions, it becomes time consuming 
when applied to a large number of functions since one need to tag functions by hand,
which clearly not tractable for large codes like numpy. Moreover, the user has most of the time no clues
about where numerical instability comes from and must therefore all functions as potentially suspect. 
This issue enforces the need to use automatic approaches.
It's worth mentioning that DBI based tools can be used for Python because they are language agnostic but
they are not intended to analyze Python code and do not take into account high-level information about 
types being used or detailed stack trace. Moreover, few tools embed a visualizer for 
analyzing numerical quality over time letting the postprocessing to the user.

\tristan{The content of this section is excellent but the structure is a bit messy. Would you like to restructure in 3 paragraphs as follows? (1) There's no numerical instrumentation tools in Python (your last paragraph + anteater), (2) DBI tools could be used but won't provide a detailed stack trace, (3) tools for other languages.}

\subsection{Detecting instabilities}

We review here the existing theoretical framework that might be used to detect numerical instabilities.

\label{sec:mca}
\subsubsection{Monte Carlo Arithmetic}
\tristan{Add a sentence to connect this part to the previous one. Something like: We use Monte-Carlo arithmetic to detect numerical instabilities. You could also summarize what the other options would be.} \tristan{cite \cite{parker1997monte}}

MCA simulates round-off errors by introducing random perturbations in floating point computations. Each floating point number is replaced by its MCA counterparts:
\[
inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the basis, $t$ the virtual precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ a random variable.
Virtual precision is one strength of the MCA method since it allows simulating 
reduced working precision.
MCA comes with three modes: Random Rounding (RR), Precision Bounding (PB) and FullMCA where either the output, either the inputs or both are perturbed with the $inexact$ function.
While the RR mode is equivalent to stochastic rounding, PB mode allows identifying 
catastrophic cancellations.

MCA provides a formula to compute the number of significant digits:
\[
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|}
\]
where $\sigma$ and $\mu$ are respectively the empirical standard deviation and expected value of a variable $x$ evaluating several times with MCA. 
Note that this formula gives an approximation of the real $s$ and 
one can refer to~\cite{sohier2018confidence} to obtain associated rigorous confidence intervals.

@@

SHAMAN, CADNATrace;
Veritracer~\cite{chatelain2018veritracer} is an extension of Verificarlo~\cite{verificarlo} that aims 
at visualizing the numerical stability over time for C, C++ and Fortran codes.

\section{PyTracer design}

@@

PyTracer follows a three step approach. First, selected functions are automatically
instrumented before executing the targeted application. During the execution, instrumented
functions will dump their inputs and outputs in a file called a trace. 
The idea is to profile several times the application to measure discrepancies between runs.
Based on the set of traces obtained from multiple executions, Pytracer 
will gather them into one trace containing statistics about each traced variables.
With all information gathered, the user can visualize and interact with it through a Ploty/Dash server.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/workflow.pdf}
    \caption{Pytracer workflow \tristan{Excellent figure but fuzzy should appear}}
    \label{fig:workflow}
\end{figure}

\subsection{Instrumentation}

Pytracer dynamically instruments the Python modules selected by the user
without requiring modifications of the targeted application.
To do so, Pytracer creates a new version of each selected module 
by preserving the attributes' signature, allowing to use them
transparently in the original application.
If by default Pytracer instruments all attributes present
in a given module, except the special attributes of the form \texttt{\_\_*\_\_},
the wrapping is not always be feasible as detailed in section ?.
The original attributes are then preserved and the user is warned.
The user can also restrict the list of attributes to trace through
a mechanism of inclusion/exclusion.

The main advantage of this method is its transparency for the user
since this latter only needs to select which modules to trace and Pytracer instrument them for him.
Moreover, the code does not need to be changed contrary to the method based on decorators, allowing
to use instrument functions when the source code is not available or editable, like for instance
in singularity containers or when running on HPC cluster.

\subsection{Module import interception}

Modules in Python are loaded through the import mechanism by the \texttt{import} keyword.
During this process, Python is looking for module path file in the default system paths and those provided by
the user (with \texttt{PYTHONPATH}). When the selected module is found, Python executes it for triggering 
instructions in the \texttt{\_\_init\_\_.py} file and stores it in the \texttt{sys.modules} map,
acting as a cache.

Pytracer intercepts the module loaded by using the \texttt{Finder} and \texttt{Loader} mechanism.
The Finder is responsible for finding the loader for a module that is being imported while the 
Loader loads and initializes the module. We add a new Finder and Loader on top of the default ones
to intercept imports and to wrap modules selected by the user. This mechanism allows returning 
the original module in the wrapper code and returning the wrapped one in the targeted application
without modifying the actual code.

\subsubsection{Module functions}

The module function is the simplest case to deal with.
For each function encountered in a module, Pytracer
creates a wrapper around this function that is dynamically compiled with the \textit{compile}
builtin function. The original function is then replaced by its wrapped counterpart.

However, this technique cannot be applied for all functions, especially for 
callable instances, which are instances of class that can be called like traditional functions
but that contain attributes. Indeed, using a wrapper would 1) modify the type of the instance
that could cause semantic bugs 2) hide attributes that causes \texttt{AttributeError} exceptions.

\begin{figure}
    \centering
\begin{lstlisting}[language=Python,style=customPython]
def getwrapperfunction(function_module, function_name, function, function_wrapper_name):
    function_wrapper_name = f
    function_id = id(function)
    info = (function_id, function_module, function_name)
    wrapper_code = f"""
    def {function_wrapper_name}(*args, **kwargs):
        return generic_wrapper({info},*args,**kwargs)"""
    return wrapper_code
\end{lstlisting}
    \caption{Caption}
    \label{fig:wrapper_code}
\end{figure}

\subsubsection{Classes}

Overriding Python classes is similar to module functions.
Pytracer goes through the attributes and replaces each callable attribute (method) by its wrapped counterpart.
The resulting class has the same type as the original one, avoiding type conflicts.
However, some classes forbid overwriting attributes of their instances, in particular the `\_\_call\_\_` operator
allowing to call an instance as a function.
Moreover, no general solution can be provided since some Python classes are non-inheritable 
like \texttt{bool},\texttt{NoneType} or \textit{universal functions} in numpy (\texttt{ufunc})
and thus cannot be inherited to overload operators.
The solution consisting in replacing function by a wrapper cannot be applied here.

This issue was mostly encountered with numpy ufunc that are intensively used in numpy. Universal functions are vectorized element-wise operators on
multidimensional arrays (\texttt{ndarrays}). They are implemented in C code with target specialized versions to achieve best performance.
One should then implement subclassing mechanism in C, which is contrary to Pytracer philosophy to be as non-intrusive and portable as possible.
However, since \texttt{ufunc} are intensively used in scientific computational codes (elementary mathematical functions are ufunc in numpy),
we implement a special treatment for them. Numpy provides the \texttt{frompyfunc} helper function to create \texttt{ufunc} from arbitrary function.
The main problem with this approach is that the output type is \texttt{PyObject} representing abstract object that conflicts 
when used as index. Figure illustrates this issue. As a last resort, user may disable \texttt{ufunc} tracing by Pytracer.

\begin{figure*}
\begin{minipage}[t]{0.4\linewidth}
    \begin{lstlisting}[language=Python,style=customPython]
>>> import numpy as np
>>> x = np.array(range(10))
>>> index = np.array([1,-1],dtype=int)
>>> x[index]
>>> array([1, 9])
    \end{lstlisting}
    \label{fig:wrapper_code}
\end{minipage}
\begin{minipage}[t]{0.4\linewidth}
    \begin{lstlisting}[language=Python,style=customPython]
>>> import numpy as np
>>> x = np.array(range(10))
>>> index = np.array([1,-1], dtype=object)
>>> x[index]
Traceback (most recent call last):
  File "<stdin>", line 1, in <module>
IndexError: arrays used as indices must be of integer (or boolean) type
    \end{lstlisting}
    \label{fig:wrapper_code}
\end{minipage}
\caption{Illustration of the issue of using \texttt{frompyfunc} function to convert function to \texttt{ufunc}.}
\end{figure*}

\subsection{Detecting instabilities}

>>>>

\documentclass[11pt]{article}
\usepackage[margin=0.5in]{geometry}
\usepackage{todonotes}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pifont}
\usepackage{xcolor,colortbl}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=black,citecolor=blue,filecolor=black,urlcolor=blue}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{makecell}
\usepackage{ulem} % provides sout
\usepackage{minted} % Provides listing environment

\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:} #1\color{black}\xspace}
\newcommand{\tristanmod}[2]{\color{orange}\sout{#1} #2\color{black}\xspace}

\newcommand{\Yohan}[1]{\color{green!75!black}\textbf{Yohan:} #1\color{black}\xspace}
\newcommand{\cross}[0]{\cellcolor{red!65}\ding{53}}
\newcommand{\valid}[0]{\cellcolor{green!75!black}\ding{51}}
\newcommand{\warn}[0]{\cellcolor{orange!75}?}
\newcommand{\pytracer}[0]{PyTracer\xspace}

\usepackage{tabularx} % provides tabularx to adjust table column sizes to page size
% define thickline for use in tables
\makeatletter
\newcommand{\thickhline}{%
    \noalign {\ifnum 0=`}\fi \hrule height 1pt
    \futurelet \reserved@a \@xhline
}
\newcolumntype{"}{@{\hskip\tabcolsep\vrule width 1pt\hskip\tabcolsep}}
\makeatother

@@

\lstdefinestyle{customC}{
  belowcaptionskip=1\baselineskip,
  breaklines=true,
  xleftmargin=\parindent,
  language=C,
  showstringspaces=false,
  basicstyle=\scriptsize\ttfamily,
  keywordstyle=\bfseries\color[rgb]{0.580, 0.000, 0.827},
  %{purple!40!lightgray},
  commentstyle=\itshape\color{green!40!black},
  identifierstyle=\bfseries\color{cyan!75!black},
  stringstyle=\color{orange},
  deletekeywords={double,float},
  classoffset=1, % starting new class
  otherkeywords={double,float},
  morekeywords={double,float},
  keywordstyle=\bfseries\color{green!55!black},
  classoffset=0
}

@@

\makeatletter
\let\orig@lstnumber=\thelstnumber
\newcommand\lstsetnumber[1]{\gdef\thelstnumber{#1}}
\newcommand\lstresetnumber{\global\let\thelstnumber=\orig@lstnumber}
\makeatother

@@

The scientific Python ecosystem is a central component of scientific
analyses due to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications every year, providing a reference set of high-quality open-source core scientific libraries. Numerous domain-specific data analysis pipelines were built from this ecosystem by diverse research groups, leveraging Python's simplicity and expressivity.

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, small numerical perturbations introduced
by data noise, software and hardware updates, or parallelization lead to
substantial deviations in final results and potentially different scientific conclusions, threatening the reliability of computational analyses. To address this issue, numerical stability analyses need to be conducted systematically. However, no practical tool currently exists to conduct such analyses on Python programs.

We present PyTracer, a numerical stability analyzer for Python programs.
PyTracer adopts a dynamic approach that evaluates numerical stability using program execution traces, enabling scaling to large programs without manual intervention. In contrast, theoretical approaches based on condition number or backward error analyses require detailed modelling of the program and its numerical implementation. Similarly, static code analysis techniques such as Frama-C~\cite{cuoq2012frama}, Gappa~\cite{de2010certifying} or Flocq~\cite{boldo2011flocq} hardly scale to large codebases, particularly in dynamically-typed languages such as Python.

@@

Several tools have been and continue to be developed to assess numerical quality. Those tools can be divided in two main categories depending whether they adopt a static or a dynamic approach.
Static analysis produces theoretically-valid error bounds 
while dynamic analysis is generally more scalable although it only provides experimental estimations.
\pytracer adopts the dynamic approach, due to its scalability to large existing code bases. Dynamic approaches typically rely on the tracing of floating-point computations to detect instabilities and visualize them. The remainder of this section reviews existing approaches for each of these three steps.

% In this section, we review tools and theoretical frameworks related to \pytracer components.
% We start by reviewing state-of-the-art dynamic tools for detecting numerical instabilities.
% Then, we focus on existing theoretical approaches for perturbations-based techniques to detect numerical instabilities,
% \pytracer being agnostic to perturbation technique used. Finally, we review visualization tools 
% aiming at detecting numerical instabilities over execution time.

\label{sec:soa}
\subsection{Numerical tracing techniques}

Many tools  are dedicated to the tracing of floating-point computations in C, C++ or FORTRAN programs due to the prevalence of these languages in High-Performance Computing (HPC). 
The main tracing techniques are source-to-source and compiler based transformations. \tristan{Re-organize the following in two paragraphs, (1) source-to-source, (2) compiler-based. For each paragraph, define the technique in 1 sentence, then list the tools} 
CADNA~\cite{jezequel2008cadna} is a library for C, C++ and Fortran  implementing the
CESTAC~\cite{vignes1993stochastic} stochastic arithmetic \tristan{is it source-to-source or compiler-based?}.
Verificarlo~\cite{verificarlo} is a compiler that supports different types of arithmetic instrumentations, including
Monte Carlo Arithmetic (MCA)~\cite{parker1997monte}. The work in ~\cite{bao2013fly} modified the GNU Compiler Collection (GCC) to track local floating-point errors across executions. pLiner~\cite{guo2020pliner} is a root cause analyzer based on the Clang compiler that detects floating-point operations 
responsible for result variability. 
pLiner uses source-to-source transformation at the AST \tristan{expand acronym} level  to rewrite parts of code with higher precision. PFPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel}, is a compiler that uses parallel shadow execution to detect numerical issues using higher precision.
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model to propagate numerical error. 
FLiT~\cite{sawaya2017flit} is framework to detect variability induced by compilers and their optimizations.
The work in~\cite{tang2016software} proposed a source-to-source framework for executing targeted code 
in infinite and fixed precision with and without using stochastic arithmetic.
The work in~\cite{wang2012development} proposes a numerical debugger based on GDB~\cite{stallman1988debugging} for 
Discrete Stochastic Arithmetic (DSA) on FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna}
and Shaman propose a GDB based tool to respectively use the CADNA and Shaman libraries with GDB.

Another technique, Dynamic Binary Instrumentation (DBI), evaluates numerical stability at the binary level,
enabling the analysis of any binary regardless of the programming language. However, working at binary level makes it difficult to access high-level information needed
for debugging and understanding the source code logic. In particular, analyzing Python code with DBI is feasible but provides very limited insights on the stability of specific Python code sections.
CraftHPC~\cite{lam2013dynamic} uses DBI to detect catastrophic cancellations at runtime.
Verrou~\cite{fevotte2016verrou} is a Valgrind~\cite{nethercote2007valgrind} based tool that replaces on-the-fly
floating-point operations by their stochastic arithmetic counterparts. FPDebug~\cite{benz2012dynamic} uses DBI to detect numerical 
inaccuracies by using shadow computations with a higher precision.
Herbgring~\cite{sanchez2017finding} is a Valgrind-based tool to detect
numerical instabilities. It uses a shadow memory to do detect precision losses by comparison with results obtained with the high-precision 
MPFR library~\cite{fousse2007mpfr}. It is combined with symbolic computation to backtrack the root of the error.
We can note that all methods based on a high-precision oracle suppose that computing with
a large number of bits is sufficient to obtain an accurate result, which is not always true, see for example the famous Muller's sequence ~\cite{chatelain2018veritracer}.

To the best of our knowledge, there is currently no tool dedicated to numerical analysis of Python code. Existing tools for tracing Python code are dedicated to performance profiling, for time (cProfile) 
or memory consumption (memprofile). 
Anteater~\cite{faust2019anteater} is a Python tracer tool that aims at debugging Python application. 
It operates source transformation at AST level but only deals with native numeric Python types.
Moreover, the user needs to manually tag each variable to trace. Finally, according to the authors,
Anteater do not scale to large traces.
However, we can compare the methodology used by these tools \tristan{are you referring to Anteater here? if so, it should be "this" tool. Otherwise, list which tools you're referring to.} against the one used in Pytracer.
These tools use an instrumentation based on decorators, a Python mechanism allowing 
to instrument a function by adding a line over a function declaration.
While this method is useful when targeting specific functions, it becomes time consuming 
when applied to a large number of functions since one needs to tag functions manually,
which is not feasible for large code bases where potentially unstable code sections are not known.

% Moreover, the user has most of the time no clues
% about where numerical instability comes from and must therefore all functions as potentially suspect. 
% This issue enforces the need to use automatic approaches.
%  Moreover, few tools embed a visualizer for 
% analyzing numerical quality over time letting the postprocessing to the user.

% \tristan{The content of this section is excellent but the structure is a bit messy. Would you like to restructure in 3 paragraphs as follows? (1) There's no numerical instrumentation tools in Python (your last paragraph + anteater), (2) DBI tools could be used but won't provide a detailed stack trace, (3) tools for other languages.}

\label{sec:detecting-instabilities}
\subsection{Detecting numerical instabilities}

Three main approaches exist to detect numerical instabilities once floating-point computations are traced: stochastic arithmetic, uncertainty or sensitivity analysis, and random seed analysis. The main techniques for each approach are reviewed below.

% We review here the existing theoretical framework that might be used to detect numerical instabilities.
% Our experiments rely on the Monte Carlo Arithmetic (MCA), a stochastic arithmetic allowing 
% to detect numerical errors by adding random perturbations in the floating-point computations.
% However, Pytracer do not rely on any assumption about the perturbation model used and thus leaves the user free to use
% an alternative model. As such, we describe two others options that we thought might be beneficial and complementary to MCA: 
% input data uncertainty analysis and random seed sensitivity.

\label{sec:mca}
\subsubsection{Stochastic Arithmetic}

% The Monte Carlo Arithmetic is a variant of so-called stochastic arithmetic, whose purpose is to model round-off errors by a random variable.

Stochastic arithmetic is a theoretical model that exploits randomness to estimate the floating-point format's numerical instabilities. The idea is to treat round-off errors as a random variable for then using classical statistical tools to study them.

The first published complete stochastic arithmetic framework is CESTAC~\cite{vignes1993stochastic}. Each floating-point computation is run multiple times in this method, switching the rounding mode to either round-up or round-down with
an equal probability. From these stochastic samples, one can then derive the number of significant digits in the computed result. In addition, the Discrete Stochastic Arithmetic (DSA~\cite{vignes2004discrete}) extends the CESTAC method to comparison operators implemented in the CADNA library.

The Monte Carlo Arithmetic works on the same principle but introduces two differences:
a virtual precision parameter allowing to simulate reduced working precision and different \texttt{modes} to introduce perturbations
on the inputs or the output.
MCA simulates round-off errors by introducing random perturbations in floating-point computations and replaces each floating-point number with its stochastic counterpart:
\[
inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the basis, $t$ the virtual precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ a random variable.
Virtual precision is one strength of the MCA method since it allows simulating reduced working precision.
MCA comes with three modes: Random Rounding (RR), Precision Bounding (PB), and FullMCA, which perturb either the output, the inputs, or both with the $inexact$ function. While the RR mode is equivalent to stochastic rounding, PB mode allows identifying catastrophic cancellations.

MCA provides a formula \tristan{is this formula specific to MCA or can it also be applied to CESTAC?} \Yohan{Yes, CESTAC supposes that the random variable is normal and so use a different formulation} to compute the number of significant digits:
\begin{equation}
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\sigma$ and $\mu$ are respectively the empirical standard deviation and expected value of a variable $x$ evaluating 
several times with MCA. 
Note that this formula gives an approximation of the real. Recently, Sohier et al. ~\cite{sohier2018confidence} introduced a 
general theoretical framework that includes DSA and MCA formulae and allows to obtain associated rigorous confidence intervals on the number of significant digits estimated.

\subsubsection{Uncertainty and sensitivity analysis}

Uncertainty analysis (UA) is a method to derive the output distribution response from an input distribution.
The idea is to model uncertainty on data due to limited precision measurement for example by a probability distribution and 
to estimate the impact on the result. 
Sensitivity analysis (SA) is quite close to uncertainty analysis but differs from the scale observed.
While uncertainty analysis attempts to describe the whole output distribution, sensitivity analysis 
measure the local response to a small perturbation~\cite{loucks2017water}. 
Several approaches for UA and SA exist among them differential analysis, response surface methodology, Monte Carlo analysis
and variance decomposition (see~\cite{helton2006survey}). In particular, Monte Carlo Analysis is a sampling-based technique
that samples inputs space to estimate output response. Monte Carlo analysis shares similarities with Monte Carlo Arithmetic. For instance, the PRECISE framework~\cite{chaitin1996lectures} shares with MCA the idea of varying virtual precision to estimate output sensitivity.

\subsubsection{Random seed sensitivity}

The quality of a Random Number Generator is a key characteristic when dealing with stochastic model and differ from cryptography requirements.
For numerical simulation a RNG of quality must have a good tradeoff between performance, period, correlation or uniformity~\cite{james2020review}.
Poor quality RNGs may have a substantial impact on results~\cite{click2011quality}. As a first step, measuring variation induced by seed is 
a good indicator about the program stability that can be compared against variability induced by floating-point arithmetic.
\Yohan{This section lacks of strong arguments. Needs improvement}

@@

Visualization covers a wide range of specific domains in the numerical simulation area, from general purpose numerical visualization with tools
like VTK~\cite{schroeder2000visualizing},
to topological data analysis~\cite{tierny2018topological}, including matrix~\cite{wu2008matrix} or tensor~\cite{kindlmann2006diffusion} visualization.
We review here tools dedicated to the analysis of the numerical variability induced by the floating-point model.
We can first highlight the fact that despite the large number of existing tools dedicated to pinpoint 
source code instructions, few of them allow the user to have a global view of the numerical accuracy of their computations
or compare several executions of the same program under various perturbations.
% Indeed, all methods based on an high-precision oracles cannot capture variability between several executions as they only refer 
% to one oracle. In other words, the actual computation is compared against the same computation with higher-precision by 
% using the same inputs data in a synchronous way.

Veritracer~\cite{chatelain2018veritracer} was the first tool to propose an embedded visualizer to 
visualize the numerical stability over time for C, C++ and Fortran codes.
It is a trace-based tool built as an extension of Verificarlo~\cite{verificarlo} that offers contextual 
information about variables traced to allow the user to quickly identify instabilities.
However, it does not offer a dynamic interface neither takes into account information about 
high-level structures like \texttt{struct} or \texttt{array}.
FPAvizual~\cite{gu2014fpavisual} is a tool for visualizing effects of floating-point arithmetic but only for educational purposes.
Anteater~\cite{faust2019anteater} proposes a visualizer for Python codes but as mentioned previously,
it only deals with native Python types and the user needs to manually select variables to trace. 
However, Anteater provides a real reflection around visual debugging beyond floating-point 
analysis, which has inspired the design of \pytracer's visualization interface.

\section{\pytracer design}

@@

The PyTracer workflow follows a three step approach (Figure~\ref{fig:workflow}). When a Python application is loaded, \pytracer instruments it using ``monkey patching", a technique to modify Python programs without editing their source code. The instrumentation replaces all the functions with a wrapper that saves their arguments and outputs in a trace file using ``pickle", Python's object serialization mechanism. \pytracer then aggregates trace files obtained from multiple perturbed executions, computing summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

% \tristan{mention monkey patching}

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figure/workflow.pdf}
    \caption{Pytracer workflow \tristan{move the execution and noise injection to step 2}}
    \label{fig:workflow}
\end{figure}

\subsection{Python code instrumentation}

%\tristan{``Instrumentation" should appear in Figure 1}

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the original application can transparently use it.
By default, \pytracer instruments all the attributes
in a given module \tristan{not only the functions?} \Yohan{No, you can have class definitions or instances that need to be wrapped}, except the special ones of the form \texttt{\_\_*\_\_}.
However, some attributes are not writable; in this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of attributes to trace through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches, which enable instrumentation in read-only environments such as Singularity containers or HPC clusters.

\subsubsection{Intercepting module import}

Python loads modules through an import mechanism triggered by the \texttt{import} keyword, using two objects: \texttt{finders} and \texttt{loaders}.
The \texttt{finder} is responsible for finding the package (or 
namespace\footnote{Python makes a distinction between \texttt{regular} and \texttt{namespace} packages.
A regular package is a directory that contains a \_\_init\_\_.py file and potentially subdirectories (sub-packages) 
that contain themselves a \_\_init\_\_.py file and so on recursively. 
The package hierarchy follows those of the directory. 
Namespace packages introduced in Python 3.3 (\href{https://www.python.org/dev/peps/pep-0420/}{PEP 420}) do not contain an
\_\_init\_\_.py file and allow for flexible directory structures. Hence, parts of the package can be located in zip files, on the network, or in separated directories. There are no functional differences between both types of packages.}) from its fully qualified name, whether it is located on the local storage, such as standard packages installed through the pip package manager, or on a distant server.
\texttt{Finders} do not load the modules; they return a specification (\texttt{spec} object) encapsulating 
information on where to find it and how to load it.
The \texttt{loader} creates and initializes a module object, filling the import-related attributes 
such as \texttt{\_\_spec\_\_}. 
Then, the \texttt{loader} executes the module and populates its namespace. Finally, the module is bound to its import name in the \texttt{sys.modules} dictionary.

Python implements default \tristanmod{\texttt{importers}}{\texttt{finder} and \texttt{loader}} classes as well as search paths for importing a module, but it also provides a way to create custom \texttt{finder} and \texttt{loader} classes.
To register a new \texttt{finder} class, one needs to add it to the \texttt{sys.meta\_path} list.
When an \texttt{import} statement is encountered, Python first looks for a binding in the \texttt{sys.modules}
and then iterates over \texttt{sys.meta\_path}.
Python queries each \texttt{finder} class successively until it finds the module or it reaches the end of the list, raising a \texttt{ModuleNotFoundError}. \pytracer adds  a custom \texttt{finder} class at the head of the list that intercepts
the module import and creates the module with a custom \texttt{loader} class.

\pytracer's \texttt{loader} class first loads the original module, then copies it as a new instance of the \texttt{ModuleType} class.
It then calls the appropriate instrumentation function for each module attribute depending on its type (function, class or submodule). Once all the attributes are instrumented, 
\pytracer replaces the original module by its instrumented counterpart
in the \texttt{sys.modules} map \tristan{is the non-instrumented version ever added to sys.modules? If so, could this lead to race conditions? If not, replace with ``adds the instrumented module to the sys.modules map"}. Hence, the calling convention remains unchanged and the application will transparently call 
the instrumented module.

% \pytracer intercepts the modules loaded with the \texttt{Finder} and \texttt{Loader} mechanisms \tristan{is it restrictive? Are there other mechanisms?}.
% The Finder searches the loader of a module that is being imported while the 
% Loader loads and initializes the module. \pytracer adds \tristan{overrides?} a new Finder and Loader on top of the default ones
% to intercept imports and to instrument modules selected by the user. This mechanism allows returning 
% the original module in the instrumentation code and returning the instrumented module in the targeted application
% without modifying its actual code.

\subsubsection{Instrumenting module functions}
